---
title: "Data Wrangling:Transforming 1/3"
author: "ynakahashi"
date: "2017/7/29"
output: html_notebook
---

### Exercise 1
Print out the hp column using the select function.

```{r cars01}
mtcars %>% select(hp) 
```

### Exercise 2
Print out the all but hp column using the select function.

```{r cars02}
mtcars %>% select(-hp)
```

### Exercise 3
Print out the mpg, hp, vs, am, gear columns. Consider using the colon (:) symbol.

```{r cars03}
mtcars %>% select(mpg, hp, vs:gear)
```

### Exercise 4
Create the object cars_m_h containing the columns mpg, hp columns but let the column names be ‘miles_per_gallon’, and ‘horse_power’ respectively.

```{r cars04}
cars_m_h <- mtcars %>% 
   select("miles_per_gallon" = mpg, "horse_power" = hp)
cars_m_h
```

### Exercise 5
Change the column names of cars_m_h from ‘miles_per_gallon’, and ‘horse_power’ to ‘mpg’ and ‘hp’ respectively.

```{r cars05}
cars_m_h <- cars_m_h %>% rename("mpg" = miles_per_gallon, "hp" = horse_power)
cars_m_h
```

### Exercise 6
Print out a randomly half the observations of cars_m_h.
Hint : consider using the sample_frac function

```{r cars06}
cars_m_h %>% sample_frac(0.5)
```

### Exercise 7
Create a cars_m_h_s object, containing from 10th to 35th row of cars_m_h.
Hint : Consider using the slice function.

```{r cars07}
cars_m_h_s <- cars_m_h %>% 
   slice(10:35)
cars_m_h_s
```

### Exercise 8
Print out the cars_m_h_s object without any duplicates.
Hint : Consider using the distinct function.

```{r cars08}
cars_m_h_s %>% distinct()
```

### Exercise 9
Print out from cars_m_h_s object all the observations which have mpg>20 and hp>100.

```{r cars09}
cars_m_h_s %>% filter(mpg > 20, hp > 100)
```

### Exercise 10
Select the ‘Lotus Europa’ car.

```{r cars10}
cars_m_h %>% filter(rownames(.) == "Lotus Europa")
```